---
title: "Workshop Instructions"
author: "Alex Skeels"
date: "`r Sys.Date()`"
format:
  html:
    css: style.css
    smooth-scroll: true
    toc: true          # generate a table of contents
    toc-location: left # place TOC in a sidebar on the left
    toc-depth: 2       # how many heading levels to include
    number-sections: false
execute:
  eval: false
  warning: false
  fig-width: 6    # inches
  fig-height: 4   # inches
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **The Stuff that's Always Here**

1.  ::: panel-tabset
    ## Learning objectives {.active}

    -   Work with spatial data: points and polygons

    -   Visualise spatial data on maps

    -   Make decisions about cleaning data

    -   Produce simple species distribution models

    ## Ground rules

    Here are a few ground rules to keep things running smoothly:

    1)  Do your best to attempt each task on your own before asking for help from the instructors or your neighbour

    2)  When you get to a green box, stick the [green]{style="color:green;"} Post-it on your laptop where we can see it

    3)  If you need help, stick the [red]{style="color:red;"} Post-it on your laptop

    ## Getting un-stuck

    Remember, no programmers know it all, so don’t struggle alone too long, but before you ask for help do try:

    -   googling for what you want to do (e.g. ‘set point transparency in ggplot’)

    -   googling the error message if you get one (try putting it in double quotes)

    -   searching stack overflow for what you want try using tags (e.g. \[ggplot\]) to narrow down your search, e.g. search \[ggplot2\] adjust point transparency

    -   use the amazing R studio cheat sheets

    -   ask ChatGPT or Gemini for help (they are good at R!)

    -   asking your neighbours if they can help (remember to be generous with your time if someone asks you!)

    -   try explaining to an imaginary person exactly what you’re trying to do

    -   try working from the examples in the help files (e.g. type ?geom_point in the console and look down to the examples)

    If none of this works, use your [red]{style="color:red;"} Post-it to get help from a demonstrator.

    ## Before you begin

    Do this at the start of the workshop. For the git commits in this part, you do not need to have descriptions (though you can add them if you prefer).

    1.  On github.com create a new private repo called DS4B_Week_N, where N is the week of the semester (e.g. DS4B_Week_4)

    2.  Add the GitHub user DS4B-ANU (https://github.com/DS4B-ANU) as a collaborator using these instructions.

    3.  Clone your repository to your computer using GitHub Desktop. The new folder on your computer is what is known as the ‘root’ directory of your repository.

    4.  Create a .gitignore file as follows:

        -   In RStudio, use the dialog box in the top left to bring up the menu to add files

        -   Choose “Text File”

        -   Add the line .Rproj.user/ to your text file

        -   Add other lines for files you need to ignore (e.g. you might like to add .DS_Store if you use a mac)

        -   Save the text file in the root directory of your repo, and call it .gitignore

        -   Commit and push your changes using GitHub Desktop

    5.  Create a new R project as follows:

        -   Open RStudio

        -   Create a new R project for this workshop using the dialog button at the top left

        -   Choose “Existing Folder” when the dialog comes up

        -   Choose the root directory of your new repo as the folder in which to create your R project (the folder will be DS4B_Week_N as above, where N is the week of the semester)

        -   Create the R project Commit and push your changes using GitHub Desktop

    6.  Now we add the raw data as follows:

        -   Create a folder in the root directory of your repo called data

        -   Download the raw data for this workshop from Canvas

        -   Save it in the data folder

        -   Commit and push your changes using GitHub Desktop.

    7.  Now we create the R Markdown file as follows:

        -   In RStudio, use the dialog box in the top left to bring up the menu to add files

        -   Choose “R Markdown”

        -   Give it the title workshop_solutions.rmd

        -   Save it in the root directory of your repo

        -   Delete the contents of the R markdown file

        -   Replace the contents with the following markdown template:

            ````
            ---
            title: "BIOL3207 Workshop Solutions"
            author: "UXXXXXXX"
            date: "YYYY-MM-DD"
            output: html_document
            ---

            ```{r setup, include=FALSE}
            knitr::opts_chunk$set(echo = TRUE)
            ```

            ## Introduction

            ## Background Points

            ## Fit Models

            ## Model Validation

            ## Model Predictions Through Time
            ````

        -   Enter your U number and the date where indicated

        -   Knit your R markdown file to html in RStudio

        -   Check the html file works by opening it in your browser

        -   Commit and push your changes using GitHub Desktop.

    8.  Open your repo on github.com, and ensure that:

        -   You have added DS4B-ANU (https://github.com/DS4B-ANU) as a collaborator

        -   Your root directory contains:

            -   a .gitignore file (containing at least the single line .Rproj.user/)

            -   a data/ folder

            -   the data files downloaded from Canvas in the data/ folder

            -   a .Rproj file with the same name as the repo (e.g. Practice_1.Rproj)

            -   a workshop_solutions.rmd file, updated as above

            -   a workshop_solutions.html file, knitted from the .rmd file

            -   nothing else!

        -   If any of these need fixing, fix them then use GitHub Desktop to commit and push your changes.

    ## Remember it’s a lab book

    Through this and every workshop, you should type as much descriptive text into your R Markdown document as someone (including but not limited to you) would need to understand what you’re doing.

    Remember: This document is your lab book! The information in it must be sufficient to reproduce your research, allow others to understand your way of thinking, and allow you to use it in your later work (e.g. when writing assessments, theses, blogs or papers). Imagine that you are writing a manual and a report at the same time.

    Please keep the level 1 headings we have given you above, so it’s easy for us to come around and help you. Please use subheadings to divide up your work within each section, subheadings work like this:

    ```{r}
    ## This is a level 2 subheading
    ### This is a level 3 subheading
    ```
    :::

## Introduction

Before you start, first make sure you’ve set up your github repo correctly following the instructions in the “Before You Begin” section above.

Today we are going to continue our journey of modelling the distribution of the Frilled Lizard (*Chlamydosaurus kingii*) with one of the most widely used tools in ecology: species distribution models (SDMs). Last week, we used exclusively geometric models which rely on presence-only data to model geographic range polygons. This week, we will use a class of SDMs that uses both presence and absence (or pseudo-absence) data to model suitable habitat based on environmental conditions.

These models underpin:

-   conservation planning\
-   invasive species forecasts\
-   climate change projections\
-   biodiversity policy

They are powerful tools but very easy to misuse. Throughout today’s workshop, focus less on *running models* and more on *understanding assumptions*.

### The Data

We are going to use the cleaned occurrence records and the associated alpha hull that we worked on last week. You may have slightly different data to your neighbours depending on how you cleaned the data and defined the alpha value for your alpha hulls. That's ok!

We will also use bioclimatic data from WorldClim, which are commonly used for SDMs which describe variation in monthly averages of temperature and precipitation, e.g., mean temperature of the warmest quarter or precipitation seasonality. More details on the definition of these variables is [here](https://www.worldclim.org/data/bioclim.html).

We begin with the cropped occurrence dataset you created last week and the alpha hull.

```{r}
library(tidyverse)
library(terra)
library(sf)
library(raster)  # needed for leaflet raster display

# occurence points
occ <- read_csv("data/frilled_lizard_ALA_cropped.csv")

# load alpha hull (saved at the end of last week's workshop)
ahull_v <- vect("data/frilled_lizard_alpha_hull.shp")

# convert to sf for use with leaflet later
ahull_sf <- st_as_sf(ahull_v)
```

Environmental SDMs require spatial layers describing climate or habitat. We will use **bioclim variables** derived from temperature and rainfall.

Load the raster stack with the `rast` function from the `terra` package.

```{r}
bioclim_data <- rast("data/worldclim_australia.tif")
```

#### **Task 1. Inspect the Predictors (10 minutes)**

Raster data can be organised as a matrix, with each row representation a latitude, and each column a longitude. It can also be organised with each row representing a unique grid cell on the landscape. This second format is useful when you are looking at multiple variables simultaneously and is what is returned by `as_tibble`.

It is useful to inspect the raw data numerically and visually. Spend a few minutes using `skimr` to look at the moments of the data, `leaflet` to visualise the data on a map.

-   Take a look at a few variables

-   Can you guess what units the variables are in?

-   Do variable ranges look reasonable?

-   Do any of variables appear correlated?\

```{r}
library(skimr)
library(leaflet)
library(leafsync)

values(bioclim_data) |>
  as_tibble() |>
  skim()

bio01 <- leaflet() |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addRasterImage(raster::raster(bioclim_data$Bio01),
                 opacity = 0.8) |>
  addLegend(pal = colorNumeric("viridis",
                              values(bioclim_data$Bio01),
                              na.color = "transparent"),
            values = values(bioclim_data$Bio01),
            title = "Bio01")


bio12 <- leaflet() |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addRasterImage(raster::raster(bioclim_data$Bio12),
                 opacity = 0.8) |>
  addLegend(pal = colorNumeric("viridis",
                              values(bioclim_data$Bio12),
                              na.color = "transparent"),
            values = values(bioclim_data$Bio12),
            title = "Bio12")

sync(bio01, bio12)
```

### Extracting Environmental Values

To model a niche, we need predictor values at occurrence locations. First, however, we want to drop points which are duplicated in a grid cell. This removes some bias from having some grid cells with many observations (e.g., close to cities) and some with very few. All grid cells with a sighting will contribute just once to our models.

```{r}
# prune data to one point per grid cell
# Create SpatVector
occ_v <- vect(occ, geom = c("X", "Y"), crs = crs(bioclim_data))

# Get raster cell IDs
occ_v$cell_id <- cellFromXY(bioclim_data, crds(occ_v))

# Keep one point per raster cell
occ_pruned <- occ_v[!duplicated(occ_v$cell_id), ]

# Extract climate
occ_env <- terra::extract(
  bioclim_data,
  occ_pruned,
  ID = FALSE
) |> 
  as_tibble()

occ_model <- bind_cols(
  as_tibble(crds(occ_pruned)) |> 
    setNames(c("longitude", "latitude")),
  occ_env
) |> 
  mutate(presence = 1)
```

::: callout-tip
## Reminder: Knit, Commit, Push!
:::

## Background Points

As mentioned above, environmental model-based SDMs require both presence and absence data. For most species, we do not have reliable absence data available so we must use what is known as background (or pseudo-absence) data.

Background data can be sampled in many ways, each will affect your results. A good study on selecting background points is [here](https://doi.org/10.1111/j.2041-210X.2011.00172.x).

Today we are going to test this directly.

### **Form Groups**

With your neighbours, split into groups of **four**.

Each member of the group will generate 5000 background points using a **different strategy**, then you will compare the resulting suitability maps.

By the end of this exercise, your group should be able to answer:

-   Do modelling decisions strongly influence predictions?
-   Which method seems most biologically realistic?
-   Which method would you trust for conservation decisions?

There is no universally correct answer — the goal is to develop modelling judgement.

#### Task 2. Generate Background Points Using Your Assigned Method (~20 minutes)

-   Within your group, assign one method to each person.

-   Run **only your tab**.

-   Afterwards, you will compare results using synced maps.

::: panel-tabset
## Method 1 {.active}

### Random Background

This is the simplest and most commonly used approach. Here, we sample environmental space uniformly across the study region.

`set.seed` can be used to set the seed of the random number generator. When drawing samples, this will ensure if you use the same seed (e.g. 123) you will always draw the same sample of points. Good for reproducibility.

```{r}
set.seed(123)

bg <- spatSample(
  bioclim_data,
  size = 5000,
  method = "random",
  na.rm = TRUE,
  as.points = TRUE
)

bg_env <- terra::extract(
  bioclim_data,
  bg,
  ID = FALSE
) |>
  as_tibble()

bg_model <- bind_cols(
  st_coordinates(st_as_sf(bg)) |>
    as_tibble() |>
    rename(longitude = X, latitude = Y),
  bg_env
) |>
  mutate(presence = 0)
```

Use `leaflet` to plot a raster + the occurrence points + the background points. What patterns do you notice?

Here is a template showing how to combine a raster layer, occurrence points, and background points on a single leaflet map. Adapt this for your method:

```{r}
leaflet() |>
  addProviderTiles(providers$OpenStreetMap.Mapnik) |>

  # Add a climate raster layer (e.g. Bio01)
  addRasterImage(raster::raster(bioclim_data$Bio01), opacity = 0.5) |>

  # Plot background points in grey
  addCircleMarkers(
    data = bg_model,
    lng = ~longitude, lat = ~latitude,
    radius = 3,
    fillOpacity = 0.4,
    stroke = FALSE,
    color = "grey"
  ) |>

  # Plot occurrence points in blue on top
  addCircleMarkers(
    data = occ_model,
    lng = ~longitude, lat = ~latitude,
    radius = 5,
    fillOpacity = 0.7,
    stroke = FALSE,
    color = "#2C7FB8"
  )
```

## Method 2

### Accessible Area Background

In this method, we sample points within a fixed distance of our occurrence points, using a buffer, just like in last weeks workshop. Here, we are assuming that the species could realistically disperse only within some distance of known occurrences. Further, we are also limiting the range of environments we will contrast with our occurrence points, which increase the specificity of the model to our species.

We buffer occurrence points to define an **accessible area**.

```{r}
library(sf)

# Convert occurrence data to sf object (needed for spatial operations)
occ_sf <- st_as_sf(occ, coords = c("X", "Y"), crs = 4326)

occ_proj <- st_transform(occ_sf, 3577)

buffer <- st_buffer(occ_proj, dist = 300000) |>
  st_union()

buffer_vect <- vect(st_transform(buffer, 4326))

bg <- spatSample(
  mask(bioclim_data, buffer_vect),
  size = 5000,
  method = "random",
  na.rm = TRUE,
  as.points = TRUE
)

bg_env <- terra::extract(bioclim_data, bg, ID = FALSE) |>
  as_tibble()

bg_model <- bind_cols(
  st_coordinates(st_as_sf(bg)) |>
    as_tibble() |>
    rename(longitude = X, latitude = Y),
  bg_env
) |>
  mutate(presence = 0)
```

Use `leaflet` to plot a raster + the occurrence points + the background points using the template from Method 1 as a guide. What patterns do you notice?

## Method 3

### Environmental Envelope Background

In this method, we sample points within Australia that are within the broad environmental range of our occurrence points. Here, we are assuming that the species could realistically disperse only within the climatic conditions that it could experience or have access to across its distribution. If its range is determined by climate, then it should only occupy a subset of the total range of conditions in the area. Further, we are also limiting the range of environments we will contrast with our occurrence points which should increase the specificity of the model to our species.

In other words, instead of restricting sampling background points in geographic space, we restrict **environmental space**.

We sample only within the environmental range + one standard deviation observed at presence locations.

```{r}
env_ranges <- occ_env |>
  summarise(
    across(
      everything(),
      list(min = ~min(.x, na.rm = TRUE)-sd(.x, na.rm = TRUE),
           max = ~max(.x, na.rm = TRUE)+sd(.x, na.rm = TRUE))
    )
  )

clim_df <- terra::as.data.frame(
  bioclim_data,
  xy = TRUE,
  na.rm = TRUE
)

ranges_long <- env_ranges |>
  pivot_longer(
    everything(),
    names_to = c("variable", ".value"),
    names_sep = "_(?=[^_]+$)"   # splits on final underscore only
  )

filtered_cells <- clim_df |>

  pivot_longer(
    cols = -c(x, y),
    names_to = "variable",
    values_to = "value"
  ) |>

  left_join(ranges_long, by = "variable") |>

  filter(value >= min,
         value <= max) |>

  group_by(x, y) |>
  summarise(n = n(), .groups = "drop") |>

  filter(n == nrow(ranges_long)) |>

  select(x, y)

bg <- filtered_cells |>
  slice_sample(n = 5000)

# Convert to SpatVector for terra::extract
bg_vect <- vect(as.data.frame(bg), geom = c("x", "y"), crs = crs(bioclim_data))
bg_env <- terra::extract(bioclim_data, bg_vect, ID = FALSE) |>
  as_tibble()

bg_model <- bind_cols(
  bg |>
    rename(longitude = x, latitude = y),
  bg_env
) |>
  mutate(presence = 0)
```

Use `leaflet` to plot a raster + the occurrence points + the background points using the template from Method 1 as a guide. What patterns do you notice?

## Method 4

### Bias Correction Background

In species distribution modeling, the choice of background points (locations representing where a species could potentially occur) can strongly influence the model. One way to make these background points more realistic is to account for sampling effort — that is, where people have actually looked for the species.

Here, we use a bias-correction approach based on the distribution of Agamid lizards in Australia:

-   The Frilled Lizard (Chlamydosaurus kingii) belongs to the family Agamidae, which includes over 70 species in Australia.

-   If a researcher is out recording Agamid lizards, it is very likely that they would record the Frilled Lizard if it were present.

-   Therefore, areas where Agamid lizards have been observed represent locations that have actually been sampled, and are reasonable places to draw background points.

-   By sampling background points from these areas, we avoid including locations where no one has ever looked, which could otherwise bias the model by suggesting that absence is equivalent to unsurveyed locations.

In other words, this method uses existing survey effort as a guide to produce more realistic background points, improving the reliability of our species distribution models.

```{r}
# Read in distribution of all agamid species
agamid_occ <- read.csv("data/agamidae.csv")

# Convert to terra vector object
agamid_occ <- terra::vect(
  agamid_occ,
  geom = c("decimalLongitude", "decimalLatitude"),
  crs = "EPSG:4326"
)

# Rasterize occurrence counts onto first climate layer
# This gives a "density" surface of sampling effort
agamid_density <- terra::rasterize(
  x = agamid_occ,
  y = bioclim_data[[1]],
  fun = "sum"
)

# Sample random points proportional to density ("bias correction")
# We normalise the density raster to create sampling probabilities,
# then use spatSample with method = "weights" so areas with more
# Agamid records are sampled more heavily.
set.seed(123)  # for reproducibility

# Normalise density to 0-1 range for use as sampling weights
agamid_density[is.na(agamid_density)] <- 0
agamid_weights <- agamid_density / max(values(agamid_density), na.rm = TRUE)

# Mask to match bioclim extent (only sample where we have climate data)
agamid_weights <- mask(agamid_weights, bioclim_data[[1]])

bck_bias <- terra::spatSample(
  x = agamid_weights,
  size = 5000,
  method = "weights",
  as.points = TRUE,
  na.rm = TRUE
)

# Extract coordinates for tidyverse manipulation
bg <- as_tibble(terra::crds(bck_bias)) |>
  rename(longitude = x, latitude = y)

# Extract environmental values for each point
bg_vect <- vect(bg, geom = c("longitude", "latitude"), crs = crs(bioclim_data))
bg_env <- terra::extract(bioclim_data, bg_vect, ID = FALSE) |>
  as_tibble()

# Combine into modeling-ready dataframe
bg_model <- bind_cols(
  bg,
  bg_env
) |>
  mutate(presence = 0)
```

Use `leaflet` to plot a raster + the occurrence points + the background points. What patterns do you notice?
:::

::: callout-tip
## Reminder: Knit, Commit, Push!
:::

## Fit Models

Now that we have **presence and background data**, along with the environmental variables extracted at each site, we can move on to the **modeling part** of the workshop.

First lets combine our occurrence and background data into a single data frame.

```{r}
sdm_data <- bind_rows(occ_model, bg_model)

# Check the result
glimpse(sdm_data)
```

We will fit two types of models:

-   **Generalized Linear Model (GLM):** Models linear relationships between predictors and the response (presence/absence). Here we use logistic regression.\
-   **Generalized Additive Model (GAM):** Can capture **non-linear relationships** using smoothing functions (`s()`), allowing the model to flexibly fit more complex patterns.

> ⚠️ Note: In real-world SDMs, models are typically fitted at high spatial resolution with many predictors and more advanced statistical or machine learning techniques. Here, we use a simplified workflow to focus on the **fundamentals**.

------------------------------------------------------------------------

#### **Task 3. Select predictors (~15 minutes)**

Selecting environmental predictors is critical. Variables should be chosen based on **ecological relevance** and to **avoid collinearity**. Both GLMs and GAMs do poorly when predictors are highly correlated.

```{r}
# Check for collinearity among predictors
library(caret)

correlation_matrix <- cor(sdm_data[, 3:21], use="complete.obs")
correlated <- findCorrelation(correlation_matrix, cutoff = 0.7)  # remove variables with |r| > 0.7
names_correlated <- colnames(sdm_data[,3:21])[correlated]

# Remove highly correlated variables
sdm_data_uncor <-sdm_data[, !colnames(sdm_data) %in% names_correlated]

# Keep track of predictors
predictor_vars <- colnames(sdm_data_uncor)[grep("Bio", colnames(sdm_data_uncor))]
print(predictor_vars)
```

**Ask yourself:**

-   Which variables were removed due to colinearity?

-   Do the remaining predictors make ecological sense for the Frilled Lizard?

-   Would you remove any others based on your ecological knowledge of the species (if any)?

Next, we define **formulas** for GLM and GAM models. For the GAM, `k = 4` sets the flexibility of the splines: increase `k` for more complex curves, decrease for simpler curves.

```{r}

# GLM: linear model
formula_glm <- formula(paste0("presence ~ ", paste0(predictor_vars, collapse = " + ")))

# GAM: non-linear model with splines (k = basis dimension, controls wiggliness)
formula_gam <- formula(paste0("presence ~ ", paste0("s(", predictor_vars,", k = 4)", collapse = " + ")))

# Inspect formulas
print(formula_glm)
print(formula_gam)
```

#### **Task 4. Fit models (~20 minutes)**

Now we fit both models using the uncorrelated predictors.

```{r}
library(mgcv)

# Fit GLM (logistic regression)
glm_sdm <- glm(formula_glm, data = sdm_data_uncor, family = binomial(link = "logit"))
summary(glm_sdm)

# Fit GAM (using mgcv, which correctly handles smooth terms in predict)
gam_sdm <- mgcv::gam(formula_gam, data = sdm_data_uncor, family = binomial(link = "logit"))
summary(gam_sdm)

```

::: callout-note
## Reading model output
In the `summary()` output, look for:

-   **Significant variables**: look for small p-values (marked with `*`, `**`, or `***`) next to each predictor.
-   **AIC**: printed at the bottom of the GLM summary. Lower AIC = better model fit.
-   **Explained deviance (D²)**: calculate as `1 - (residual deviance / null deviance)` from the summary output. Higher D² = more variance explained. In R: `1 - (glm_sdm$deviance / glm_sdm$null.deviance)`. For the GAM, `summary(gam_sdm)` prints "Deviance explained" directly.
:::

**Ask yourself:**

-   Which variables are significant in each model?

-   Which model has a lower AIC (better fit)?

-   Which model has a higher explained deviance (D²)?

Now we project the niche into geographic space by using `predict` to apply the fitted model to the climate data for the whole of Australia. For every grid cell, the model predicts the relative habitat suitability (a value between 0 and 1).

```{r}
pred_glm <- terra::predict(bioclim_data,
                       glm_sdm,
                       type = "response")

pred_gam <- terra::predict(bioclim_data,
                       gam_sdm,
                       type = "response")

# Create a color palette
pal <- colorNumeric(
  palette = "Spectral",
  domain = c(0, 1),
  na.color = "transparent"
)

glm_plot <- leaflet() |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addRasterImage(raster::raster(pred_glm), colors=pal, opacity = 0.8) |>
  addLegend(
    pal = pal,
    values = values(pred_glm),
    title = "GLM Suitability",
    position = "bottomright"
  )

gam_plot <- leaflet() |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addRasterImage(raster::raster(pred_gam), colors=pal, opacity = 0.8) |>
  addLegend(
    pal = pal,
    values = values(pred_gam),
    title = "GAM Suitability",
    position = "bottomright"
  )

sync(glm_plot, gam_plot)

```

We can turn a continuous measure of habitat suitability into a binary species distribution by thresholding the predictions. To do this we need to evaluate our models and assess how well the model predicts our known occurence records.

A common approach is the **Receiver Operating Characteristic (ROC) curve**, which plots the trade-off between **true positive rate** and **false positive rate** across all possible thresholds.

From the ROC curve, we calculate the **Area Under the Curve (AUC)**:

-   **AUC ≈ 1** → the model perfectly separates presences from absences

-   **AUC ≈ 0.5** → the model performs no better than random guessing

::: callout-warning
## Note
Higher AUC values indicate better predictive performance, but AUC alone does not capture all aspects of model reliability — it should be interpreted alongside other evaluation metrics. Also note that because we are using pseudo-absences (background points) rather than true absences, the predicted values represent **relative habitat suitability** rather than true probability of occurrence.
:::

```{r}
library(dismo)

# Split presence and background
p <- sdm_data_uncor[sdm_data_uncor$presence==1, ]
a <- sdm_data_uncor[sdm_data_uncor$presence==0, ]

# Evaluate models
# evaluate() returns an object with slots accessed using @ (e.g. glm_eval@auc)
# This is different from $ which you use for data frames — @ is for S4 objects
glm_eval <- evaluate(p=p, a=a, model=glm_sdm, type='response')
gam_eval <- evaluate(p=p, a=a, model=gam_sdm, type='response')

# Plot ROC curves
par(mfrow=c(1,2))
plot(glm_eval, "ROC")
plot(gam_eval, "ROC")

```

Take a look at the evaluation objects and try plotting some other summary statistics, like `kappa`, the true positive rate (`TPR`), false positive rate `FPR`, etc.

The aim is to find the value of habitat suitability, which has the maximum value of a metric like `spec_sens` - the threshold at which the sum of the sensitivity (true positive rate) and specificity (true negative rate) is highest. Then we say all sites with a habitat suitability higher than that threshold is classed as presence, and every cell with a habitat suitability lower than the threshold is classed as an absence.

```{r}
pred_glm_threshold <- threshold(glm_eval, stat = "spec_sens")
pred_gam_threshold <- threshold(gam_eval, stat = "spec_sens")

pred_glm_bin <- pred_glm > pred_glm_threshold
pred_gam_bin <- pred_gam > pred_gam_threshold


glm_bin_plot <- leaflet() |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addRasterImage(raster::raster(pred_glm_bin), opacity = 0.8)

gam_bin_plot <- leaflet() |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addRasterImage(raster::raster(pred_gam_bin), opacity = 0.8)

sync(glm_bin_plot, gam_bin_plot)
```

Now, use `leaflet` to add the species occurrence points and the alpha hull onto this plot.

**Ask yourself:**

-   How well do these align?

-   Are there occurrence points in areas deemed unsuitable?

-   Are the suitable areas very far from any known occurrences?

::: callout-tip
## Reminder: Knit, Commit, Push!
:::

## Model Validation

So far we have evaluated the model on the same data used to fit it, so it may appear to perform very well. This is not a realistic estimate of predictive performance, because in real-world applications we want the model to generalize to new locations.

A better way to estimate predictive performance is to split the data into training and testing subsets, fit the model on the training data, and evaluate it on the testing data.

#### **Task 5. K-Fold Cross-Validation (~15 minutes)**

K-fold cross-validation automates the validation process by splitting the data into `k` partitions and repeats the evaluation on each partition. As follows:

1.  Split the data into K folds

2.  Repeat K times:

    -   Train the model on K – 1 folds

    -   Test the model on the remaining fold

3.  Combine results across folds to get an average AUC and measure of variability

```{r}
K <- 5
k_fold <- kfold(sdm_data_uncor, k = K)

# Set up table to store results
auc_df <- data.frame(k = 1:K, glm_AUC = NA, gam_AUC = NA)

for (k in 1:K) {
  
  # Split into training and testing data
  train <- sdm_data_uncor[k_fold != k, ]
  test  <- sdm_data_uncor[k_fold == k, ]
  
  # Refit models on training data
  mod_glm <- update(glm_sdm, data = train)
  mod_gam <- update(gam_sdm, data = train)
  
  # Evaluate on test fold
  glm_eval_tmp <- evaluate(
    p = test[test$presence == 1, ],
    a = test[test$presence == 0, ],
    model = mod_glm,
    type = 'response'
  )
  
  gam_eval_tmp <- evaluate(
    p = test[test$presence == 1, ],
    a = test[test$presence == 0, ],
    model = mod_gam,
    type = 'response'
  )
  
  # Save AUC for each fold
  auc_df[k, "glm_AUC"] <- glm_eval_tmp@auc
  auc_df[k, "gam_AUC"] <- gam_eval_tmp@auc
}

# Inspect AUC across folds
print(auc_df)
```

By repeating the evaluation on different subsets of the data, **k-fold cross-validation provides a more robust estimate of model predictive performance** than a single train/test split.

::: callout-tip
## Reminder: Knit, Commit, Push!
:::

## Model Predictions Through Time

If we're happy with our model in the present-day and convinced we have accurately modelled the niche of the species, we might want to ask how the species distribution has changed through time.

Commonly, people look to model the change in distribution into the future using models of climate change under different emissions scenarios. Equally, we might also want to know how the species was distributed in the past to understand how population structure might have evolved over time when conditions were vastly different to today.

One such period is the Last Glacial Maximum \~21,000 years ago. During the LGM, Australia looked very different, it was a colder and drier place than the present day and many species may have contracted their geographic ranges to warmer and more mesic refugia.

Lets see what our SDM predicts for the distribution of the Frilled Lizard at the LGM!

First, lets load the LGM climate data make sure the names match the WorldClim data

```{r}
# load bioclim data for the LGM (from CHELSA)
lgm_bioclim_data <- rast("data/chelsa_australia_LGM.tif")

# The names of the different variables must match exactly
names(bioclim_data)
names(lgm_bioclim_data)

# check they match — if not, force the names to align
names(bioclim_data) %in% names(lgm_bioclim_data)

# ensure names match exactly (some climate datasets use different naming)
names(lgm_bioclim_data) <- names(bioclim_data)
```

Now we have data that represents the same variables, with matching names, at two different times. We can now predict the model on this new data.

```{r}
pred_glm_lgm <- terra::predict(lgm_bioclim_data, glm_sdm, type="response")
pred_gam_lgm <- terra::predict(lgm_bioclim_data, gam_sdm, type="response")

pred_glm_bin_lgm <- pred_glm_lgm > pred_glm_threshold
pred_gam_bin_lgm <- pred_gam_lgm > pred_gam_threshold
```

Before we plot it out, lets do the same for the year 2100. There are many models for future climate than use different climatological models and shared socioeconomic pathways (SSP). WorldClim has data available for many of these - see [here](https://www.worldclim.org/data/cmip6/cmip6_clim2.5m.html).

Today we'll use MIROC6 climate model and SSP585 which is the highest emissions scenario.

```{r}
# load bioclim data for 2100 (from WorldClim MIROC6 SSP585)
future_bioclim_data <- rast("data/worldclim_australia_MIROC6_2100.tif")

# check they match — if not, force the names to align
names(bioclim_data) %in% names(future_bioclim_data)
names(future_bioclim_data) <- names(bioclim_data)

# make the model prediction
pred_glm_2100 <- terra::predict(future_bioclim_data, glm_sdm, type="response")
pred_gam_2100 <- terra::predict(future_bioclim_data, gam_sdm, type="response")

# threshold the model prediction
pred_glm_bin_2100 <- pred_glm_2100 > pred_glm_threshold
pred_gam_bin_2100 <- pred_gam_2100 > pred_gam_threshold
```

### Final Task (~20 minutes)

Each person in your group should pick the model algorithm they think performed best (GLM or GAM) and create three leaflet maps synced together using `leafsync::sync()`: one for the LGM range, one for the present-day range, and one for the 2100 distribution. On top of the present-day map, add the occurrence points and the alpha hull for comparison.

::: callout-warning
## Extrapolation caution
When projecting models to past or future climates, the environmental conditions may fall outside the range the model was trained on. This can produce unreliable predictions, especially for GAMs where smooth terms can behave unpredictably outside the training data range. Interpret temporal projections as indicative, not definitive.
:::

You can sync three maps like this: `sync(map_lgm, map_present, map_future, ncol = 1)`

To add the alpha hull to a leaflet map, use the `ahull_sf` object we created at the start: `addPolygons(data = ahull_sf, fillOpacity = 0.2, color = "red")`

Next, get together with your group and compare each member's plot and answer these questions.

-   Which background selection step gave you the best estimate of the present-day range?

-   How did the background selection methods differ from one another?

-   Can you think of a possible reason some methods performed better than others? (no wrong answers!)

::: callout-tip
## ✅ Check out!

Once you’ve completed the workshop up to this point (not including any of the Going Further section), you are welcome to check out. **But first, knit, commit, and push!**

If you’d like to stay and complete the ‘Going Further’ section, you’re welcome to do so — but please check out first.

**Before checking out, make sure you have:**

-   Fitted GLM and GAM models with cross-validation results
-   At least one suitability map (continuous or binary) plotted with leaflet
-   The group comparison of background point methods discussed
-   A three-panel temporal comparison (LGM, present, future)
-   Written reflections on the "Ask yourself" questions in your `.rmd`
-   Your `.rmd` knits to `.html` without errors

**To check out:** bring your laptop to Alex, Rob, or a demonstrator with both your `.rmd` and `.html` files open.
:::

## Going Further

### (1) Thresholding with different validation metrics

In the workshop so far, you have performed thresholding on the range with the metric `spec_sens`, but there are many other possible choices to use. read about the other options (`?threshold`) and try some of these out. How does the estimate of the range change? Which metrics predict a larger range? and which a smaller?

### (2) Cross-validation with spatial folds

In the workshop so far, we have used random folds in K-fold cross validation. These generally show consistent high performance across folds. However, to assess whether a model is transferable in space and time, it is better to split the data in a more structured way. For example, by splitting into spatial folds, we can ask whether the distribution of a species in one restricted part of its range can be informative about where it should be distributed in other parts of the continent. A robust model should be transferable. A rough script to create 6 spatial folds is below but check out the `blockCV` package for even more options to do this.

```{r}

# Create 2D spatial blocks
sdm_data_uncor <- sdm_data_uncor |>
  mutate(
    lon_bin = ntile(longitude, 3),
    lat_bin = ntile(latitude,  2),
    spatial_fold = as.numeric(interaction(lon_bin, lat_bin))
  )

# If number of folds exceeds K, randomly reduce to K
K <- length(unique(sdm_data_uncor$spatial_fold))

auc_df_spatial <- data.frame(k = 1:K, glm_AUC = NA, gam_AUC = NA)

for (k in 1:K) {
  
  # Spatial split
  train <- sdm_data_uncor[sdm_data_uncor$spatial_fold != k, ]
  test  <- sdm_data_uncor[sdm_data_uncor$spatial_fold == k, ]
  
    if (length(unique(test$presence)) < 2) {
    next
  }

  
  # Refit models
  mod_glm <- update(glm_sdm, data = train)
  mod_gam <- update(gam_sdm, data = train)
  
  # Evaluate
  glm_eval_tmp <- evaluate(
    p = test[test$presence == 1, ],
    a = test[test$presence == 0, ],
    model = mod_glm,
    type = "response"
  )
  
  gam_eval_tmp <- evaluate(
    p = test[test$presence == 1, ],
    a = test[test$presence == 0, ],
    model = mod_gam,
    type = "response"
  )
  
  auc_df_spatial[k, "glm_AUC"] <- glm_eval_tmp@auc
  auc_df_spatial[k, "gam_AUC"] <- gam_eval_tmp@auc
}

print(auc_df_spatial)

```
